Analyzing Biological Models with the Wasserstein Distance

### 2016-05-24 Paper/Paper.tex
updates

% do not change this formatting 
\documentclass[10pt]{article}
\parindent=0pt \parskip=8pt
\textwidth=3in
\hoffset=.3in 
\usepackage{amssymb,amsmath,amsthm,mathrsfs} 
\pagestyle{headings}
\usepackage{amsmath}
\newcommand*{\QEDA}{\hfill\ensuremath{\blacksquare}}%
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{url}
\usepackage{shadethm}
\usepackage{mathtools}
\usepackage{systeme}
\usepackage{float}
\usepackage{mathtools}
\setlength\parindent{24pt}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
\usepackage{caption}
\usepackage{subcaption}
\newtheorem{theorem}{Theorem}
\usepackage{dcolumn}
\newcolumntype{2}{D{.}{}{2.0}}
\usepackage{subfig}

@@

\subsection{One Dimensional Case}
For illustration, we begin by examining the growth of cancer cells. To model such a phenomenon on a macroscopic level, we study the model
\begin{equation}
\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u},
\label{eq: MacroModel}
\end{equation}
where $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. The solution to ($\ref{eq: MacroModel}$) is the well-known Gaussian and is given by

\begin{equation*}
Macro_\rho(x) = \frac{1}{\sigma \sqrt{2\pi}}\exp\Big(\frac{-(x-\mu)^2}{2\sigma^2}\Big)
\end{equation*}

\noindent For our micro-model, we consider the model given by
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel}
\end{equation}
where $B_t$ denotes a standard Brownian motion, with $B(0) = 0$. In order to compare our micro-model to the macro-model, we approximate (\ref{eq: MicroModel}). For $\Delta t > 0$, we have
\begin{equation}
\int_0^{\Delta t} \ dX_t = \sigma \int_{0}^{\Delta t} \ dB_t \implies x_{\Delta t} - x_0 = \sigma \textbf{Z}
\end{equation}
where $\textbf{Z} \sim \mathcal{N}(0, \Delta t)$. Using properties of variance, 
\begin{equation}
x_{\Delta t} = x_0 + \sigma \sqrt{\Delta t}\textbf{W}
\label{eq: BrownianEQ}
\end{equation} 
where $\textbf{W} \sim \mathcal{N}(0, 1)$. To compute $Micro_{\rho}$, $N \in \mathbb{Z}^+$ i.i.d. realizations of (\ref{eq: BrownianEQ}) are computed, from which a histogram ($Micro_{\rho}$) is created over the physical domain.  In Figures \ref{fig: rhoInitial} and \ref{fig: rhoFinal} we plot $Macro_{\rho}, Micro_{\rho}$ at the initial and final times, respectively.
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{InitialDistributions.jpg}
  \caption{Solutions at initial time}
  \label{fig: rhoInitial}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{FinalDistribution.jpg}
  \caption{Solutions at final time}
  \label{fig: rhoFinal}
\end{subfigure}
\caption{$N = 10^3, \ T_{final} = 1, \ \Delta t =.01.$}
\label{fig:test}
\end{figure}

\noindent Using a standard normal PDF as the initial condition for $Macro_{\rho}$, and $\Delta x = .05$, we see in Figures \ref{fig:FittingError} and \ref{fig: FittingLogError} the decay of WD and log(WD) against number of particles, respectively.  All WD reported are at $T_{final} = 1, \ \Delta t = .01.$

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{WDVSN1D.jpg}
  \caption{Fit WD and N}
  \label{fig:FittingError}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{LogWDVSN1D.jpg}
  \caption{Fit Log(WD) and Log(N)}
  \label{fig: FittingLogError}
\end{subfigure}
\caption{Decay of WD with N}
\label{fig:test}
\end{figure}

\noindent We now explore how changing $\Delta x$, for a fixed $N$, changes the WD.

\begin{figure}[H]
\begin{center}
\includegraphics[scale = .4]{WDVSDX.jpg}
\end{center}
\caption{WD vs $\Delta x$.}
\end{figure}

\noindent We conclude from Figure \ref{fig:FittingError}  that the error between Macro$_\rho$ and Micro$_\rho$ follows the relationship 
\begin{equation*}
\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
\end{equation*}
 where $C \approx 3.33, \ \alpha \approx 0.44 $. When computing the linear fit  $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$. We thus conjecture that the Wasserstein distance between Micro$_\rho$ and Macro$_\rho$ decays at a rate of $1/\sqrt{N}$.
\begin{theorem}
$\Big|W^2_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
\end{theorem}
\begin{proof}
Denote Macro$_\rho$ by $\rho_{mi}$, Micro$_\rho$ by $\rho_{ma}$. Then, for some $\phi \in C_b(\mathbb{R})$
\begin{equation*}
\langle \rho_{mi}, \phi \rangle = \int_{\mathbb{R}} \rho_{mi} \phi(x) \ dx = \frac{1}{N}\sum_x \phi(x_i(t)).
\end{equation*}
By the Law of Large Numbers, 
\begin{equation*}
\frac{1}{N}\sum_x \phi(x_i(t)) \xrightarrow{N \rightarrow \infty} \int_{\mathbb{R}} \phi(x)\rho_{mi}(x) \ dx + \frac{1}{\sqrt{N}}\text{std}(\phi(x))
\end{equation*}
where $\text{std}(\phi(x))$ denotes the standard deviation of $\phi$.
\end{proof}

\noindent We now consider a model with a source term, meaning the macro-level model is given by

\begin{equation}
\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 u} + \rho(1-\rho),
\label{eq: MacroModel2}
\end{equation}

To calculate $\rho$Micro, $N$ realizations of (\ref{eq: BrownianEQ}) are computed. At each time-step, a new particle is created with probability $\Delta t(1-\rho)$ and destroyed with probability . A histogram ($Micro_{\rho}$) is then 
computed over the physical domain.

@@

In two dimensions, the equation for the macro model becomes 
\begin{equation}
\frac{\partial \rho}{\partial t} = D\Big(\frac{\partial^2 \rho}{\partial^2 x} + \frac{\partial^2 \rho}{\partial^2 y}\Big) + \rho(1-\rho),
\label{eq: Macro2D}
\end{equation}
where $\rho(t,x,y)$ denotes the density of cancer cell at time $t$ at location $(x,y).$ Similarly, (\ref{eq: BrownianEQ}) becomes 
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel2}
\end{equation}
where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.

>>>>

% do not change this formatting 
\documentclass[10pt]{article}
\parindent=0pt \parskip=8pt
\textwidth=3in
\hoffset=.3in 
\usepackage{amssymb,amsmath,amsthm,mathrsfs} 
\pagestyle{headings}
\usepackage{amsmath}
\newcommand*{\QEDA}{\hfill\ensuremath{\blacksquare}}%
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{url}
\usepackage{shadethm}
\usepackage{mathtools}
\usepackage{systeme}
\usepackage{float}
\usepackage{mathtools}
\setlength\parindent{24pt}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
\usepackage{caption}
\usepackage{subcaption}
\newtheorem{theorem}{Theorem}
\usepackage{dcolumn}
\newcolumntype{2}{D{.}{}{2.0}}
\usepackage{subfig}
\usepackage{mwe}

@@

\subsection{One Dimensional Case}
For illustration, we begin by examining the growth of cancer cells. To model such a phenomenon on a macroscopic level, we study the model
\begin{equation}
\frac{\partial \rho}{\partial t}  = \frac{\sigma^2}{2}\frac{\partial^2 \rho}{\partial^2 x} + \rho(1-\rho),
\label{eq: MacroModel}
\end{equation}
where $\sigma$ is a scalar and $\rho(x,t)$ denotes the density of cancer cell at time $t$ and a location $x$. For the micro-model, we consider the equation
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel}
\end{equation}
where $B_t$ denotes a standard Brownian process, with $B(0) = 0$. For $\Delta t > 0$, we have
\begin{equation}
\int_0^{\Delta t} \ dX_t = \sigma \int_{0}^{\Delta t} \ dB_t \implies x_{\Delta t} - x_0 = \sigma \textbf{Z}
\end{equation}
where $\textbf{Z} \sim \mathcal{N}(0, \Delta t)$. Using properties of variance, 
\begin{equation}
x_{\Delta t} = x_0 + \sigma \sqrt{\Delta t}\textbf{W}
\label{eq: BrownianEQ}
\end{equation} 
where $\textbf{W} \sim \mathcal{N}(0, 1)$. To compute $Micro_{\rho}$, $N \in \mathbb{Z}^+$ i.i.d. realizations of (\ref{eq: BrownianEQ}) are computed, from which a histogram ($Micro_{\rho}$) is created over the physical domain.  In Figures \ref{fig: rhoInitial} and \ref{fig: rhoFinal} we plot $Macro_{\rho}, Micro_{\rho}$ at the initial and final times, respectively.
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{InitialDistributions.jpg}
  \caption{Solutions at initial time}
  \label{fig: rhoInitial}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{FinalDistribution.jpg}
  \caption{Solutions at final time}
  \label{fig: rhoFinal}
\end{subfigure}
\caption{$N = 10^3, \ T_{final} = 1, \ \Delta t =.01.$}
\label{fig:test}
\end{figure}

\noindent Using a standard normal PDF as the initial condition for $Macro_{\rho}$, and $\Delta x = .05$, we see in Figures \ref{fig:FittingError} and \ref{fig: FittingLogError} the decay of WD and log(WD) against number of particles, respectively. Figure \ref{fig: WDVsX} shows that WD decays linearly a $\Delta x$ decreases.  All WD reported are at $T_{final} = 1, \ \Delta t = .01.$

\begin{figure}
\begin{subfigure}{.5\linewidth}
\centering
\includegraphics[scale = .4]{WDVSN1D.jpg}
\caption{Fitting WD to N}
\label{fig:FittingError}
\end{subfigure}%
\begin{subfigure}{.5\linewidth}
\centering
\includegraphics[scale = .4]{LogWDVSN1D.jpg}
\caption{Fitting Log(WD) to Log(N)}
\label{fig: FittingLogError}
\end{subfigure}\\[1ex]
\begin{subfigure}{\linewidth}
\centering
\includegraphics[scale = .4]{WDVSDX.jpg}
\caption{Fitting WD to $\Delta x$}
\label{fig: WDVsX}
\end{subfigure}
\caption{One Dimensional WD Computations}
\label{fig:test}
\end{figure}

\noindent We conclude from Figure \ref{fig:FittingError}  that the error between Macro$_\rho$ and Micro$_\rho$ follows the relationship 
\begin{equation*}
\Big|W_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\Big({\frac{1}{N}}\Big)^{\alpha},
\end{equation*}
 where $C \approx 3.33, \ \alpha \approx 0.44 $. When computing the fit $log(WD) = a\times log(N) + b$, we find $a \approx -.41,  b \approx .53$. We thus conjecture that the Wasserstein distance between Micro$_\rho$ and Macro$_\rho$ decays at a rate of $1/\sqrt{N}$.
\begin{theorem}
$\Big|W_2\Big(Macro_\rho, Micro_\rho\Big)\Big| \approx C\sqrt{\frac{1}{N}}$, for some $C \in \mathbb{R}$.
\end{theorem}
\begin{proof}
Denote Macro$_\rho$ by $\rho_{mi}$, Micro$_\rho$ by $\rho_{ma}$. Then, for some $\phi \in C_b(\mathbb{R})$
\begin{equation*}
\langle \rho_{mi}, \phi \rangle = \int_{\mathbb{R}} \rho_{mi} \phi(x) \ dx = \frac{1}{N}\sum_x \phi(x_i(t)).
\end{equation*}
By the Law of Large Numbers, 
\begin{equation*}
\frac{1}{N}\sum_x \phi(x_i(t)) \xrightarrow{N \rightarrow \infty} \int_{\mathbb{R}} \phi(x)\rho_{mi}(x) \ dx + \frac{1}{\sqrt{N}}\text{std}(\phi(x))
\end{equation*}
where $\text{std}(\phi(x))$ denotes the standard deviation of $\phi$.
\end{proof}

Figures \ref{fig: MicroVel} and \ref{fig: MacroVel} plot the respective velocities of $Micro_{\rho}, Macro_{\rho}$. To find the velocity between functions $f_1, f_2$ at time $t_j$, we set a value of $c = .1$, then find $x_{j+1} = f_{2}^{-1}(c),  x_{j} = f_{1}^{-1}(c)$. Then, the velocity at time $j$ is $v(j) = \frac{x_{j+1}- x_{j}}{\Delta t}$. Clearly, the velocity of $Micro_{\rho} \approx 1.38 \times 10^2$ is a random variable, while the velocity of $Macro_{\rho}\approx 1.5 \times 10^2$ is deterministic.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{1DMicroVelocity.jpg}
  \caption{Fitting Micro$_{\rho}$ velocity}
  \label{fig: MicroVel}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{1DMacroVelocity.jpg}
  \caption{Fitting Macro$_{\rho}$ velocity}
  \label{fig: MacroVel}
\end{subfigure}
\caption{Velocities, $T_{final} = 1, \Delta t = .01, N = 10^4, \sigma = \sqrt{2}.$}
\label{fig:test}
\end{figure}

We now investigate how the ratio $\frac{N}{\Delta x}$ impacts the WD. We consider the cases when $(N, \ \Delta x) = (10, 1), \ (10^2, .5), \ (10^3, .125), \ (10^4, .0625).$ For each case, $1,000$ i.i.id realizations of (\ref{eq: MicroModel}) are run. The WD between $Macro_{\rho}$ and $Micro_{\rho}$ at $T_{final = 1},\ \Delta t = .01$ is computed, and then the max of these WD values is computed. We plot the corresponding max WD and pairings of $(N, \ \Delta x)$ in Figures \ref{fig: FittingWDNDX} and \ref{fig: FittingLogWDNDX}.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{WDVSNDX.jpg}
  \caption{Fitting WD to $\frac{N}{\Delta x}$}
  \label{fig: FittingWDNDX}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{LogWDVSNDX.jpg}
  \caption{Fitting Log(WD) to $Log(\frac{N}{\Delta x})$}
  \label{fig: FittingLogWDNDX}
\end{subfigure}
\caption{One Dimensional WD Computations}
\label{fig:test}
\end{figure}

@@

In two dimensions, the equation for the macro model becomes 
\begin{equation}
\frac{\partial \rho}{\partial t} = D\Big(\frac{\partial^2 \rho}{\partial^2 x} + \frac{\partial^2 \rho}{\partial^2 y}\Big) + \rho(1-\rho),
\label{eq: Macro2D}
\end{equation}
where $\rho(t,x,y)$ denotes the density of cancer cell at time $t$ at location $(x,y).$ Similarly, (\ref{eq: BrownianEQ}) becomes 
\begin{equation}
dX_t = \sigma dB_t,
\label{eq: MicroModel2}
\end{equation}
where $B_t$ denotes a standard Brownian motion in 2 dimensions, with $B(0) = (0,0)$.  Figures \ref{fig:Fitting2DMicroVelocity} and \ref{fig: Fitting2DMacroVelocity}
plot the respective velocities of $Micro_{\rho}, Macro_{\rho}$.  Clearly, the velocity of $Micro_{\rho} \approx $ is a random variable, while the velocity of $Macro_{\rho}\approx 1.5 \times 10^2$ is deterministic.

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{WDVSNDX.jpg}
  \caption{Fitting WD to $\frac{N}{\Delta x}$}
  \label{fig:Fitting2DMicroVelocity}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{2DMacroVelocity.jpg}
  \caption{Fitting Log(WD) to $Log(\frac{N}{\Delta x})$}
  \label{fig: Fitting2DMacroVelocity}
\end{subfigure}
\caption{2D Velocities, $T_{final} = 1, \Delta t = .01, N = 10^3, \sigma = \sqrt{2}.$}
\label{fig:test}
\end{figure}
